---
title: "Human ovarian cancer"
author: "Jonatan A. González"
date: '2023-05-09'
geometry: margin = 2cm
fontsize: 11pt
output: pdf_document
classoption: a4paper
header-includes:
 - \usepackage{mathrsfs}
---

```{r setup, include=FALSE, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(spatstat)
library(ggplot2)
library(sf)
library(spdep)
library(GET, quietly = T)
library(doParallel)
library(sparr)
library(lmtest)
library(gee)
```

```{r, echo=FALSE}
#Loading ovarian cancer data
load("Data/CancerData.RData")
```

# Ploting some tissue samples

Here we have a sample of a patient where we display the tumour cells and colour the different immune types.

```{r, fig.align='center', fig.width=10, fig.height=4, echo=FALSE}
# Plotting Immune cells
plot(DataCells[c(4, 11, 27, 50)], cols = 1:4, main ="", cex = 0.2, 
     which.marks = "Immune", main.panel = "", ncols = 2, mar.panel=c(0.5,0,0,0.4), 
     equal.ribbon=TRUE, equal.scales = T, legend = F, mar = c(0,0,0,0)) #7x7.3
legend("bottom", inset=.00,
       legend = levels(DataCells$`1`$marks$Immune), 
       col=1:4, horiz=TRUE, cex=0.8, pch = 1:4)

# Plotting Immune and Tissue cells
par(mar = c(0, 0, 0, 0))
plot(DataCells$`6`, main  = "", which.marks = c("Immune", "Tissue"), cex = 0.4, 
     cols = yarrr::piratepal("pony", trans = .6), pch = 16, ncols = 1, 
     leg.args= list(cex = 0.4))
```

# A regression analysis of immune cell counts

Let $n_{ij}$ denote the numbers of immune cells for the $j$th subject in the $i$th group $(i = 1,2)$. For analysing possible density changes, a Poisson log-linear model for the expected cell counts could be used taking into account the possible overdispersion of the data.

```{r, echo = FALSE}
#Function for building a dataset with patient outcomes
NpointsC <- function(PPP){
  Index <- expand.grid(Immune = levels(PPP$marks$Immune), 
                       Tissue = levels(PPP$marks$Tissue))
  fi <- function(Y, Z) {npoints(subset.ppp(PPP,
                                           (PPP$marks$Immune == Y) & 
                                             (PPP$marks$Tissue == Z)))}
  nijk <- unlist(Map(f = fi, Y = Index$Immune, Z = Index$Tissue))
  Index$nijk <- nijk
  Index$Offnijk <- npoints(PPP)
  Index$Aij <- rep(log(area(PPP)), length(nijk))
  Index$ID <- factor(rep(attributes(PPP)$"Index", length(nijk)))
  Index$primary <- factor(rep(attributes(PPP)$"primary", length(nijk)), 
                          levels = 0:1, labels = c("No","Yes"))
  Index$chemo <- factor(rep(attributes(PPP)$"chemo", length(nijk)), levels = 0:1, 
                        labels = c("No","Yes"))
  Index$stage <- factor(as.numeric(rep(attributes(PPP)$"stage", length(nijk))), 
                        levels = 1:4, labels = c("I","II", "III", "IV"))
  Index$death <- factor(rep(attributes(PPP)$"death", length(nijk)), levels = 0:1, 
                        labels = c("No","Yes"))
  Index$BRCA_mutation <- factor(rep(attributes(PPP)$"BRCA_mutation", length(nijk)), 
                         levels = 0:1, labels = c("No","Yes"))
  Index$age_diagnosis <- rep(attributes(PPP)$"age_diagnosis", length(nijk))
  Index$parpi_inhibitor <- factor(rep(attributes(PPP)$"parpi_inhibitor", length(nijk)))
  return(Index)
}

NN <- do.call('rbind', lapply(DataCells, NpointsC))
```

```{r, echo = FALSE}
#Gee model
N2 <- gee(nijk ~ Immune + Tissue + offset(log(Offnijk / Aij)) +  primary + 
            chemo + stage + BRCA_mutation + age_diagnosis + parpi_inhibitor + death, 
         data = NN, id = ID, corstr = "unstructured", family = "poisson")
summary(N2)
```

To obtain p-values, we proceed as follows

```{r}
AA <- as.data.frame(2 * pnorm(abs(coef(summary(N2))[,5]), lower.tail = FALSE))
names(AA) <- "P"
AA
```

# Multitype point patterns

In this case, we need to analyse point patterns consisting of several different types of points, and they are called multitype point patterns. Such patterns are usually presented as *marked point patterns* with categorical marks. A point process $\mathbf{X}$ is split into several sub-processes of types $1,...m,$ denoted by $\mathbf{X}^{(1)},\ldots,\mathbf{X}^{(m)}$. Every functional descriptor can be calculated for each point pattern $X^{(j)}$.

# First-order intensity functions

The first-order intensity function describes the expected value of points of some kind anywhere in the observation window. There are several techniques to calculate it; the non-parametric methods are the most used in the context of point processes. We employ an adaptive kernel estimator

```{r, eval=FALSE, echo = FALSE}
# We choose a global bandwidth through the maximal principle 
bw.h0 <- mean(sapply(DataCells, OS))
# Making a hyperframe with all the point patterns in a column
Cells <- hyperframe(pp = DataCells)
```

```{r, eval=FALSE, echo  = FALSE}
# To estimate the intensity functions we use paralelisation
intenI <- function(PPP, h0 = NULL, Marks = "Immune"){
  Index <- levels(PPP$marks[[Marks]])
  In <- mclapply(Index, function(Y) {
    densityAdaptiveKernel.ppp(subset(PPP, PPP$marks[[Marks]] == Y), h0 = h0)}, 
    mc.cores = detectCores() - 1)
  names(In) <- Index
  return(as.solist(In))
}

# Computing the intensities and saving them in the hyperframe as well
# This part is time consuming (Approx two minutes per case using 5 cores)
Cells$Lambda.I <- with(Cells, intenI(pp, h0 = bw.h0, Marks = "Immune"))
Cells$Lambda.T <- with(Cells, intenI(pp, h0 = bw.h0, Marks = "Tissue"))
Cells$Lambda.bullet <- with(Cells, densityAdaptiveKernel.ppp(pp, bw = bw.h0))
```

The next figure displays samples of the intensities of some immunity markers

```{r, fig.align='center', fig.width=10, fig.height=3, echo=FALSE}
plot((Cells$Lambda.I$`6`[c(1,2,3,4)]), #9x3
     col = hcl.colors(256, "Spectral"), 
     box = F, ncols = 4,
     mar.panel=c(0,1,1,0.4), main = "", equal.ribbon = T, panel.end = Cells$pp$`6`$window)
```

The superposition intensity is defined as the intensity of the point pattern regardless the types of points, i.e., $$
\lambda_{\bullet} = \sum_{j=i}^m \lambda_j.
$$

Therefore, we can speak of the probability distribution of types.

```{r, eval=FALSE, echo = FALSE}
# This is time consuming, approx 1 minute
Cells$P.I <- with(Cells, relrisk(unmark(pp) %mark% pp$marks$Immune, diggle = T))
```

```{r, fig.align='center', result.width='100%', fig.width=9, echo=FALSE}
par(mfrow = c(2, 1)) #9x3in
plot(Cells$P.I$`6`[1:4], col = hcl.colors(256, "Roma"), box = F, ncols = 4,
     mar.panel=c(0,1,1,0.4), main = "", equal.ribbon = T, panel.end = Cells$pp$`6`$window)
```

```{r, eval=FALSE, echo = FALSE}
Cells$Dominant <- with(Cells, im.apply(P.I, which.max))
Cells$Dominant <- with(Cells, eval.im(factor(Dominant, levels = 1:4, labels = levels(pp$marks$Immune))))
```

```{r, fig.align='center', result.width='100%', fig.width=8, echo=FALSE}
par(mar = c(0, 0, 0, 0))
textureplot(Cells$Dominant$`6`, box = F, main = "") #6.75 x 3.5
```

```{r, eval=FALSE, echo = FALSE}
# This is time consuming, approx. 3 minutes with 5 cores
Segregation <- mclapply(Cells$pp, function(x) {
  segregation.test.ppp(unmark(x) %mark% x$marks$Immune, nsim = 99, verbose = F, 
                       diggle = F, sigma = bw.h0)}, 
  mc.cores = detectCores() - 1)
p.adjust(sapply(Segregation,"[[", "p.value"), "bonferroni")
```

Now we smooth the continuous marks

```{r, eval=FALSE, echo = FALSE, warning=FALSE}
Cells$pstat3 <- with(Cells, Smooth.ppp(unmark(pp) %mark% pp$marks$pstat3, diggle = T, warn = F))
Cells$ier3 <- with(Cells, Smooth.ppp(unmark(pp) %mark% pp$marks$ier3, diggle = T, warn = F))
```

```{r, fig.align='center', result.width='100%', fig.width=9, echo=FALSE}
#par(mfrow = c(2, 1)) #9x1.9in
plot(Cells$pstat3[c(4, 11, 27, 48, 50)], col = hcl.colors(256, "Berlin"), 
     box = F, main = "",equal.ribbon = F, ncols =5,
     main.panel = "")
plot(Cells$ier3[c(4, 11, 27, 48, 50)], col = hcl.colors(256, "Tofino", rev = T), 
     box = F, main = "",equal.ribbon = F, ncols =5,
     main.panel = "")
```

For the relative risk we take a random tissue sample for illustration purposes as it is time consuming

```{r, eval=FALSE, echo = FALSE}
rrisk <- function(X){
  QQ <- unmark(X) %mark% X$marks$Immune
  QQ1 <- split(QQ)
  Combi <- combn(4, 2, simplify = F)
  
  h0 <- OS(QQ, nstar = "geometric")
  hp <- OS(QQ) / 2
  
  RR <- function(c) {risk(f = QQ1[[c[1]]], g = QQ1[[c[2]]], h0 = h0, hp = hp, 
                          adapt = T, tolerate = T, pilot.symmetry = "pooled", verbose = F)}
  RR1 <- lapply(Combi, RR)
  return(RR1)
}

RandPatient <- sample(1:51,1)

#This is time consuming, approx. 3 minutes
rrTissue <- rrisk(Cells$pp[[RandPatient]])
```

```{r, fig.align='center', result.width='100%', fig.width=9, fig.height=3,, echo=FALSE}
par(mfrow = c(1, 6), mar = c(0,0,2,2.5)) #10.5 x 1.7
for (i in 1:6) {
  plot(rrTissue[[i]], col = hcl.colors(256, "OrRd", rev = T), tol.type = "two.sided", 
     auto.axes = F, main = combn(levels(Cells$pp$`6`$marks$Immune), 2)[,i])
}

```

# Second-order analysis

## Ripley's $K$-function

### Univariate case

```{r, eval=FALSE, echo = FALSE}
Rmax <- min(with(Cells, rmax.rule("K", W = pp$window, lambda = intensity(pp))))
r0 <- seq(0, Rmax, length.out = 150)
Cells$Lfun <- with(Cells, Linhom(pp, correction = "translate", r = r0, 
                                 lambda = Lambda.bullet, unitname = "microns"))
Lfuns <- collapse.fv(Cells$Lfun, same = c("r", "theo"), different = "trans")
```

The centred $L$- functions $(L(r)-r)$ of the point patterns of immune cells look like

```{r, fig.align='center', fig.width=5, fig.height=4, echo=FALSE}
plot(Lfuns, . - r ~ r, legend = F, ylab = "Inhomogeneous centred L", main = "", lwd = 0.5)
```

### Multivariate case

We compute now the $L$-functions for Stroma vs Tumour tissues

```{r, eval = FALSE, echo = FALSE}
LpatientTissue <- function(PP1, II){
  PPP <- unmark(PP1) %mark% PP1$marks$Tissue
  EstimKCross0 <- Lcross.inhom(PPP, i = "Stroma", j = "Tumor",
                               lambdaI = II$Stroma, lambdaJ = II$Tumor,
                               correction = "translate", r = r0)

  EstimHyper <- hyperframe(Lcross = list(EstimKCross0), 
                           primary = factor(attributes(PP1)$"primary",
                                            levels = 0:1, labels = c("No", "Yes")),
                           chemo = factor(attributes(PP1)$"chemo", 
                                          levels = 0:1, labels = c("No","Yes")),
                           stage = factor(as.numeric(attributes(PP1)$"stage"), 
                                           levels = 1:4, labels = c("I","II", "III", "IV")),
                           death = factor(attributes(PP1)$"death", 
                                           levels = 0:1, labels = c("No","Yes")),
                           parpi_inhibitor = factor(attributes(PP1)$"parpi_inhibitor"),
                           debulking = factor(attributes(PP1)$"debulking"))
  return(list(LCrossTissue = EstimKCross0, HyperTissue = EstimHyper))
}

LmultiTissue <- with(Cells, LpatientTissue(pp, Lambda.T))
Cells$LmultiTissue <- lapply(LmultiTissue, '[[','LCrossTissue')
BigHyperTissue <- do.call('rbind', lapply(LmultiTissue, '[[', 'HyperTissue'))
```

The cross $L$-functions (centred by subtracting $r$) of all point patterns are displayed in the next figure,

```{r, fig.align='center', fig.width=5, fig.height=4, echo=FALSE}
AlltissuesK <- collapse.fv(Cells$LmultiTissue, same = c("r", "theo"), different = "trans")
plot(AlltissuesK, . - r ~ r, legend = F, 
     ylab = "Inhomogeneous centred cross L", main = "Tumour vs Stroma", lwd = 0.5)
```

We compute now the $L$-functions for immunity markers

```{r, eval = FALSE, echo = FALSE}
Lpatient <- function(PP1, II){
  PP <- unmark(PP1) %mark% PP1$marks$Immune
  Type <- levels(PP$marks)
  Combi <- combn(Type, 2, simplify = F)
  L.Cross <- function(Labels){
   LC <- Lcross.inhom(PP, i = Labels[1], j = Labels[2], 
                      lambdaI = II[[Labels[1]]], lambdaJ = II[[Labels[2]]],
                      correction = "translate", r = r0)
   attr(LC, "units") <- "microns"
   return(LC)
  }
  EstimKCross0 <- lapply(Combi, L.Cross)
  EstimHyper <- hyperframe(Lcross = EstimKCross0, 
                           Cells = as.factor(sapply(Combi, function(x) paste(x[1], x[2], 
                                                                             sep = "_"))),
                           chemo = factor(rep(attributes(PP1)$"chemo", length(Combi)), 
                                          levels = 0:1, labels = c("No","Yes")),
                           stage = factor(rep(as.numeric(attributes(PP1)$"stage"), 
                                              length(Combi)), 
                                           levels = 1:4, labels = c("I","II", "III", "IV")),
                           death = factor(rep(attributes(PP1)$"death", length(Combi)), 
                                           levels = 0:1, labels = c("No","Yes")),
                           parpi_inhibitor = factor(rep(attributes(PP1)$"parpi_inhibitor", 
                                                        length(Combi))))
  EstimKCross <- collapse.fv(EstimKCross0, same = c("r", "theo"), different = "trans")
  return(list(EstimKCross = EstimKCross, EstimHyper = EstimHyper, levs = Combi))
}
LmultiMarker <- with(Cells, Lpatient(pp, Lambda.I))
Cells$LmultiMarker <- lapply(LmultiMarker, '[[','EstimKCross')
BigHyperMarker <- do.call('rbind', lapply(LmultiMarker, '[[', 'EstimHyper'))
```

The cross $L$-functions (centred by subtracting $r$) of the point patterns are displayed in the next figure,

```{r, fig.align='center', result.width='80%', fig.width=9, fig.height=4, echo=FALSE}
KijC0 <- function(Level){
  collapse.fv(BigHyperMarker$Lcross[BigHyperMarker$Cells == Level], 
              same = c("r", "theo"), different = "trans")
}

L0 <- unique(BigHyperMarker$Cells)
KijC1 <- lapply(L0, KijC0)

par(mfrow = c(2, 3), oma = c(1.5, 1.5, 0, 1), mar = c(1.5,1,2,1))
for (i in 1:6) { #9x4.5in
  plot(KijC1[[i]], .-r ~ r, legend = F, #ylim = c(-300,50),
     ylab = "", main = "", col = c(1, rep(i+1, 51)), lty = c(1, rep(2, 51)), lwd = .5)
  legend("bottomleft", bty = "n", #inset=.00,
       legend = L0[i], col=i+1, horiz=T, cex=0.9, lty = 2)
}
```

```{r}
# Load J multitype inhomogeneous
source("R/Jmulti.inhom.R")
```

```{r, eval = FALSE, echo = FALSE}
Jpatient <- function(PP1, II){
  PP <- unmark(PP1) %mark% PP1$marks$Immune
  Type <- levels(PP$marks)
  Combi <- combn(Type, 2, simplify = F)
  J.Cross <- function(Labels){
    ii <- (marks(PP) == Labels[1])
    jj <- (marks(PP) == Labels[2])
    lambdai <- safelookup(II[[Labels[1]]], PP[ii])
    lambdaj <- safelookup(II[[Labels[2]]], PP[jj])
    JC <- Jmulti.inhom(PP, I = ii, J = jj, r = r0, lambdaI = lambdai, 
                       lambdaJ = lambdaj, lambdamin = min(lambdaj) / 1.1)
   attr(JC, "units") <- "microns"
   return(JC)
  }
  EstimJCross0 <- lapply(Combi, J.Cross)
  EstimHyper <- hyperframe(Jcross = EstimJCross0)
  EstimJCross <- collapse.fv(EstimJCross0, same = c("r", "theo"), different = "bord")
  return(list(EstimJCross = EstimJCross, EstimHyper = EstimHyper, levs = Combi))
}
# This could be time consuming, approx. 3 mins
JmultiMarker <- with(Cells, Jpatient(pp, Lambda.I))
Cells$JmultiMarker <- lapply(JmultiMarker, '[[','EstimJCross')
BigHyperMarkerJ <- do.call('rbind', lapply(JmultiMarker, '[[', 'EstimHyper'))
BigHyperMarker$Jcross <- BigHyperMarkerJ$Jcross
rm(BigHyperMarkerJ)
```

The cross $J$-functions are displayed in the next figure,

```{r, fig.align='center', result.width='80%', fig.width=9, fig.height=4, echo=FALSE}
JijC0 <- function(Level){
  collapse.fv(BigHyperMarker$Jcross[BigHyperMarker$Cells == Level], 
              same = c("r", "theo"), different = "bord")
}

JijC1 <- lapply(L0, JijC0)

par(mfrow = c(2, 3), oma = c(1.5, 1.5, 0, 1), mar = c(1.5,1,2,1))
for (i in 1:6) { #9x4.5in
  plot(JijC1[[i]], legend = F, #ylim = c(-300,50),
     ylab = "", main = "", col = c(1, rep(i+1, 51)), lty = c(1, rep(2, 51)), lwd = .5)
  legend("topleft", bty = "n", #inset=.00,
       legend = L0[i], col=i+1, horiz=T, cex=0.9, lty = 2)
}
```

```{r}
# independence for just a sample tissue (the third)
Ex1 <- unmark(Cells$pp$`3`)
marks(Ex1) <- data.frame(Immune = Cells$pp$`3`$marks$Immune, 
                         Li = safelookup(Cells$Lambda.I$`3`[[3]], Ex1),
                         Lj = safelookup(Cells$Lambda.I$`3`[[4]], Ex1))
II <- levels(Ex1$marks$Immune)[3]
JJ <- levels(Ex1$marks$Immune)[4]
W0 <- erosion.owin(Ex1$window, r = 350)


JI <- function(X, Ii, Jj, r = NULL, W1 = W0){
  X <- X[W1]
  li <- X$marks$Li[X$marks$Immune == Ii]
  lj <- X$marks$Lj[X$marks$Immune == Jj]
  lm <- min(li, lj) / 1.1
  Jcross.inhom(unmark(X) %mark% (X$marks$Immune), i = Ii, j = Jj, 
               lambdaI = li, lambdaJ = lj, lambdamin = lm, r=r)
}

# We reduced here the number of simulations to only 99 for illustration purposes
E34 <- envelope(Ex1, JI, funargs = list(Ii = II, Jj= JJ, r = r0),
              simulate = expression(rshift(Ex1, radius = 350, edge = "erode")), 
              nsim = 99, savefuns = T)

Eget34 <- rank_envelope(E34, type = "rank")
```

```{r, fig.align='center', result.width='80%', fig.width=9, fig.height=4, echo=FALSE}
plot(Eget34)
```

```{r}
# Random labelling
Ex1 <- unmark(Cells$pp$`3`) %mark% Cells$pp$`3`$marks$Immune
Li <- safelookup(Cells$Lambda.I$`3`[[4]], Ex1)
Ld <- safelookup(Cells$Lambda.bullet$`3`, Ex1)
II <- levels(Ex1$marks)[4]
li <- Li[Ex1$marks == II]
lm <- min(Ld) / 1.1

JR <- function(X, Ii, r = NULL){
  Jidot <- Jdot.inhom(X, i = Ii, lambdaI = li, lambdadot = Ld, lambdamin = lm, r=r)
  J <- Jinhom(unmark(X), lambda = Ld, lmin = lm, r = Jidot$r)
  dif <- eval.fv(Jidot - J)
  return(dif)
}

E4 <- envelope(Ex1, JR, funargs = list(Ii = II, r = r0),
               simulate = expression(rlabel(Ex1)), 
               nsim = 99, savefuns = T)

Rget4 <- rank_envelope(E4, type = "rank")
```

```{r, fig.align='center', result.width='80%', fig.width=9, fig.height=4, echo=FALSE}
plot(Rget4)
```

# One-way graphical functional ANOVA

## Comparing interactions between cell types

```{r, eval = FALSE, echo = FALSE}
Cells$stage <- with(Cells, attr(pp, "stage"))
Cells$chemo <- with(Cells, attr(pp, "chemo"))
Cells$death <- with(Cells, attr(pp, "death"))
Lci <- sapply(Cells$Lfun, function(x) {
  U <- as.data.frame(x)
  return(U$trans - U$theo)})
Lciget <- create_curve_set(list(r = r0, obs = Lci))
LciMu <- graph.fanova(nsim = 999, curve_set = Lciget, 
                      groups = factor(Cells$stage, levels = 1:4, labels=c("I","II","III","IV")),
                      #groups = factor(Cells$chemo, levels = 0:1, labels = c("No","Yes")),
                      #groups = factor(Cells$death, levels = 0:1, labels = c("No","Yes")),
                      #variances = "unequal", n.aver = 30, mirror = T, savefuns = F, 
                      contrasts = F, 
                      typeone = "fwer",
                      test.equality = "mean")
```

```{r, fig.align='center', result.width='80%', fig.width=9, fig.height=4, echo=FALSE}
plot(LciMu)
```
